Structure Learning for $\VAR(p)$ Models

### 2019-06-21 paper/paper.tex
More references for paper plus some prose.

In practice, Granger's notion of causation it not a convincing test
for \textit{true} causation, since our statements about causation are
highly dependent upon the data that we are able to observe.  Rather,
the interpretation of Granger causality is as a means of uncovering a
flow of ``information'' or ``energy'' through some underlying graph of
interactions.  Though this graph cannot be observed directly, we will
infer it's presence as a latent structure among our observed time
series data.  This intuition is levereged in a variety of applications
e.g. in Neuroscience as a means of recovering interactions amongst
brain regions \cite{bressler2011wiener}, \cite{anna_paper2008},
\cite{david2008identifying}; in the study of the dependence and
connectedness of financial institutions \cite{NBERw16223}; gene
expression networks \cite{Fujita2007}; and power system design
\cite{Misyrlis2016450}, \cite{yuan2014root}.

Finding the ``best'' graph structure consistent with observed data is
generally an extremely challenging problem (i.e. this is a best subset
selection problem, see \cite{bss_mio}, \cite{hastie_bss_comp}), though
the comparison of quality between different structures, and hence the
notion of ``best'', needs to be quantified.  In applications where we
are interested merely in minimizing the mean squared error of a linear
one-step-ahead predictor, then we will naturally desire an entirely
dense graph of connections, since each edge can only serve to reduce
estimation error.  However, since the number of edges scales
quadratically in $n$, it is imperative to infer a sparse causality
graph for large systems, both to avoid overfitting observed data, as
well as to aid the interpretability of the results.

@@

A second and wildly successful heuristic is the LASSO regularizer
\cite{tibshirani1996regression}, which can be understood as a natural
convex relaxation to penalizing the count of the non-zero edges.  The
LASSO enjoys strong theoretical guarantees \cite{wainwright2009sharp},
even in the case of time series data \cite{basu2015}
\cite{wong2016lasso}, variations on which have been applied elsewhere
\cite{DBLP:journals/corr/HallacPBL17} \cite{haufe2008sparse}
\cite{bolstad2011causal} \cite{he2013stationary}.

@@

@@

Under the null hypothesis $H_0$ (i.e. that $x_j \npwgc x_i$) we can
form the asymptotically $\chi^2(p_{ij})$ distributed test statistic
(\hl{TODO: Citation that this is chi2})

@@

Therefore, we can make the statement that Algorithm
\ref{alg:pwgr_heuristic} asymptotically implements Algorithm
\ref{alg:pwgr}.  \hl{TODO: Modify Algorithm 2 to make this true.}
\hl{TODO: Enclose this into a theorem}

@@

@@

    \BlankLine
    \caption{Finite Sample Pairwise Graph Recovery (PWGC)}
    \label{alg:finite_pwgc}

\end{algorithm}

@@

In reference to figure \ref{fig:simulation_results_comparison1} it
should not be overly surprising that our PWGC algorithm performs
better than the LASSO for the case of a strongly causal graph, since
in this case the assumptions which guarantee the correctness of
Algorithm \ref{alg:pwgr_heuristic} hold.  However, the performance is
still markedly superior in the case of a more general DAG.

@@

\printbibliography
\end{document}

>>>>

In practice, Granger's notion of causation it not a convincing test
for \textit{true} causation, since our statements about causation are
highly dependent upon the data that we are able to observe.  Rather,
the interpretation of Granger causality is as a means of uncovering a plausible
flow of ``information'' or ``energy'' through some underlying graph of
interactions.  Though this graph cannot be observed directly, we will
infer it's presence as a latent structure among our observed time
series data.  This intuition is levereged in a variety of applications
e.g. in Neuroscience as a means of recovering interactions amongst
brain regions \cite{bressler2011wiener}, \cite{anna_paper2008},
\cite{david2008identifying}; in the study of the dependence and
connectedness of financial institutions \cite{NBERw16223}; gene
expression networks \cite{Fujita2007},
\cite{grouped_graphical_granger_modelling_for_gene_expression_regulatory_networks_discovery},
\cite{discovering_graphical_Granger_causality_using_the_truncating_lasso_penalty},
and power system design \cite{Misyrlis2016450}, \cite{yuan2014root}.

\hl{TODO: From the gene network papers, it looks like these gene
  networks may actually have strongly-causal structure.  i.e. we can
  highlight the fact that our deeper understanding of a special
  problem case may in fact have significant applicability in these
  application areas.}

Other citations: support recovery consistency for the weighted lasso
in univariate AR(p) models:

Finding the ``best'' graph structure consistent with observed data is
generally an extremely challenging problem (i.e. this is a best subset
selection problem, see \cite{bss_mio}, \cite{hastie_bss_comp}), though
the comparison of quality between different structures, and hence the
notion of ``best'' needs to be quantified.  In applications where we
are interested merely in minimizing the mean squared error of a linear
one-step-ahead predictor, then we will naturally desire an entirely
dense graph of connections, since each edge can only serve to reduce
estimation error.  However, since the number of edges scales
quadratically in $n$, it is imperative to infer a sparse causality
graph for large systems, both to avoid overfitting observed data, as
well as to aid the interpretability of the results.

@@

A second and wildly successful heuristic is the LASSO regularizer
\cite{tibshirani1996regression}, which can be understood as a natural
convex relaxation to penalizing the count of the non-zero edges.  The
LASSO enjoys fairly strong theoretical guarantees
\cite{wainwright2009sharp}, extending largely to the case of
stationary time series data with a sufficiently fast rate of
dependence decay \cite{basu2015} \cite{wong2016lasso}
\cite{autoregressive_process_modelling_via_the_lasso_procedure}.
Variations on the LASSO have been applied in a number of different
contexts \cite{DBLP:journals/corr/HallacPBL17} \cite{haufe2008sparse}
\cite{bolstad2011causal} \cite{he2013stationary}
\cite{grouped_graphical_granger_modelling_for_gene_expression_regulatory_networks_discovery}.
One of the key improvements to the original LASSO algorithm is the
adaptive (i.e. weighted) adaLASSO \cite{adaptive_lasso_zou2006}, for
which oracle results are established.  The crux of the matter for time
series analysis is to craft a good weighting scheme, see e.g.
\cite{discovering_graphical_Granger_causality_using_the_truncating_lasso_penalty}.
Success of the LASSO has lead to the famous ``bet on sparsity''
principle -- that is, for large problems and finite sources of data,
you might as well bet that a sparse model is correct, because if it
isn't, there is no hope.

In the context of time series data, sparsity assumptions remain
important, but there is significant additional structure that may
arise as a result of considering the topology of the underlying
Granger-causality graph, which to our knowledge remains largely
unexplored.  The focus of this paper is to shed light on some of these
topological questions, in particular, we study a particularly simple
notion of causality graph topology which we term ``strongly causal''
and show that $VAR$ models having this structure satisfy naturaal
intuitive notions of ``information flow'' through the graph.
Moreover, we show that such graphs are perfectly (asymptotically)
recoverable with only \textit{pairwise} Granger-causality tests.  Our
finite sample results are based on simulations, where we show
promising results for these graphs topologies where our algorithm
performs substantially better in our simulation setup than do
competing LASSO algorithms, even for graphs that do not exactly
satisfy our strongly-causal topology assumptions.

\todo{What graphs in practice may be strongly causal -- see the gene network papers.}

@@

\todo{Simulations with very small T (i.e. T = 20)}

@@

Under the null hypothesis $H_0$ (i.e. that $x_j \npwgc x_i$) we can
form the asymptotically $\chi^2(p_{ij})$ distributed test statistic
\todo{Is this actually $\chi^2$ when $p_{ij}$ is chosen from the data?  This is adaptive!}

@@

Therefore, we can make the statement that Algorithm
\ref{alg:finite_pwgc} asymptotically implements Algorithm
\ref{alg:pwgr}.  \hl{TODO: Modify Algorithm 2 to make this true.}
\hl{TODO: Enclose this into a theorem}

@@

After the graph topology $\widehat{\gcg}$ has been estimated via
Algorithm \ref{alg:finite_pwgc}, we refit the entire model with the
specified sparsity pattern directly via ordinary least squares.
Producing graph estimates which are not strongly causal can be
achieved out by performing sequential estimates
$\widehat{x}_1(t), \widehat{x}_2(t), \ldots$ by estimating a strongly
causal graph with the residuals of the previous model as input, and
then refitting on the estimated sparsity pattern.  However we reserve
any analysis of this scheme for future work.

@@

    \BlankLine
    \caption{Finite Sample Pairwise Graph Recovery (PWGC)}
    \label{alg:finite_pwgc}
    \Input{Estimates of pairwise Granger-causality statistics $F_{ij}$
      (eqn \ref{eqn:gc_statistics}).  Matrix of edge probabilities $P_{ij}$ (eqn \ref{eqn:edge_inclusion_probability}).  Hypothesis testing threshold $\delta$ chosen via the Benjamini-Hochberg criterion (Section \ref{sec:error_rate_control})}
    \Output{A strongly causal graph $\widehat{\gcg}$}
    \Initialize{$S = [n]$  \texttt{\# unprocessed nodes}\\
      $E = \emptyset$  \texttt{\# edges of }$\widehat{\gcg}$\\
      $k = 1$ \texttt{\# a counter used only for notation}}

    \BlankLine

    $W_\delta \leftarrow \{(i, j)\ |\ P_{ji} > 1 - \delta, F_{ji} > F_{ij}\}$  \texttt{\# candidate edges}\\
    $\mathcal{I}_0 \leftarrow \big(\sum_{j\in S: (j, i) \in W_\delta} P_{ij}, \mathsf{\ for\ }i \in S \big)$ \texttt{\# total node incident probability}\\
    $P_0 \leftarrow \{i \in S\ |\ \mathcal{I}_0(i) < \ceil{\text{min}(\mathcal{I}_0)}\}$ \texttt{\# Nodes with fewest incident edges}\\
    \If{$P_0 = \emptyset$}{
      $P_0 \leftarrow \{i \in S\ |\ \mathcal{I}_0(i) \le \ceil{\text{min}(\mathcal{I}_0)}\}$ \texttt{\# Ensure non-empty}
    }
    \BlankLine

    \While{$S \ne \emptyset$}{
      $S \leftarrow S \setminus P_{k - 1}$ \texttt{\# remove processed nodes}\\
      % $\mathcal{I}_k \leftarrow \big(\sum_{j \in S: (j, i) \in W_\delta} F_{ij}, \mathsf{\ for\ }i \in S \big)$ \texttt{\# intra-}$S$ \texttt{incident strength}\\
      $\mathcal{I}_k \leftarrow \big(\sum_{j\in S: (j, i) \in W_\delta} P_{ij}, \mathsf{\ for\ }i \in S \big)$\\
      $P_k \leftarrow \{i \in S\ |\ \mathcal{I}_k(i) < \ceil{\text{min}(\mathcal{I}_k)}\}$\\
      \If{$P_k = \emptyset$}{
        $P_k \leftarrow \{i \in S\ |\ \mathcal{I}_k(i) \le \ceil{\text{min}(\mathcal{I}_k)}\}$
      }
      \;
      \texttt{\# add strongest edges, maintaining strong causality}\\
      $U_k \leftarrow \bigcup_{r = 1}^k P_{k - r}$ \texttt{\# Include all forward edges}\\
      \For{$(i, j) \in \mathsf{sort}\Big(\{(i, j) \in U_k \times P_k\ |\ (i, j) \in W_\delta\} \mathsf{\ by\ descending\ } F_{ji}\Big)$} {
        \If{$\mathsf{is\_strongly\_causal}(E \cup \{(i, j)\})$} {
          \texttt{\# }$\mathsf{is\_strongly\_causal}$ \texttt{can be implemented by keeping}\\
          \texttt{\# track of ancestor / descendant relationships}\\
          $E \leftarrow E \cup \{(i, j)\}$
        }
      }
      $k \leftarrow k + 1$\\
    }
    \Return{$([n], E)$}
\end{algorithm}

@@

In reference to figure \ref{fig:simulation_results_comparison1} it
should not be overly surprising that our PWGC algorithm performs
better than the LASSO for the case of a strongly causal graph, since
in this case the assumptions which guarantee the correctness of
Algorithm \ref{alg:finite_pwgc} hold.  However, the performance is
still markedly superior in the case of a more general DAG.

@@

\clearpage
\printbibliography
\end{document}
